---
title: "Flow Cytometry Data Analysis Tutorial"
author: "bioinformatics.ca with modifications by M.L. Larsen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
fontsize: 11pt
geometry: margin = 0.75in
---

#Overview

Flow cytometry data is often complex and can be difficult to analyze without the appropriate tools and pipeline.
In this tutorial, you will learn how to import FCS data, create custom gates, and process the information per event to evaluate sample variation using a data set collect with an HIV experiment.

**This documentation was generated from the bioinformatics.cs [workshop](http://bioinformatics.ca/workshops/2013/flow-cytometry-data-analysis-using-r-2013#material) on processing flow cytometry data**

##Document workflow
###Work environment

###Module 2: Visual data exploration
###Module 3: 1D static gating
###Module 4: 1D Staitc gating
###Module 5: 1D Dynamic gating
###Module 6: Clustering and Additional FCM tools
###Module 7: Example: Analyzing DNA/RNA content in bacterial cells
####7.1 Data collection and methodology
####7.2 Explore data
####7.3 Create gates for analysis

**Flow Cytometry BioConductor Packages**

**Flow Cytometry Dictionary**

| Term  | Abbreviation  | Decription |
|:-:    |:-:            |:-:|
| FCS   |               |   |
|       | SSC           |   |
|       | FSC           |   |
|       | FS-PMT        |   |
| Compensation  |               |   |

____

Set up your local work environment and load the packages needed for the data analysis

```{r,echo=FALSE}
sessionInfo()$R.version$version.string
packinfo <- installed.packages(fields = c("Package","Version"))
#packinfo["",c("Package","Version")]
```

```{r, echo=FALSE, results='hide', message = FALSE, warning = FALSE}
rm(list = ls())
getwd()

#setwd("~/GitHub/flow-cytometry")
setwd("C:/Users/Megan/GitHub/flow-cytometry")

# load source code 
source("http://bioconductor.org/biocLite.R")
#source("supportCode/flowPrep.R")
#source("supportCode/support_functions.R")

# Install bioconductor packages. This process will take a bit of time, so be patient.
#biocLite(c("flowPeaks","flowCore","flowStats","flowViz",
#           "flowClust","flowQ", "flowUtils","flowMeans","flowDensity"))
#biocLite("GEOmap");biocLite("Logicle")
biocLite("flowQ")

# Install and load packages, source code, and dependencies
library(flowPeaks) 
library(flowCore)
library(flowStats) 
library(flowViz) 
library(flowQ) 
library(flowClust)
library(flowUtils) 
library(flowMeans)
library(flowDensity)
library(GEOmap)
#library(Logicle)

```

Each bioconductor package comes with a vignette that provides a task-oriented description of the functionality. Let's take a look at one.

```{r, results=hide}
browseVignettes(package = "flowMeans")
```

# Module 2: Exploring FCM data
For this tutorial, we will be using a data set collected from an HIV study. [data source] (http://flowrepository.org/id/FR-FCM-ZZZK)
Let's start by importing our tutorial data file. 

```{r}
# Import data as a flowFrame object which includes the meta data collected from the analyzer

f <- read.FCS("./tutorial/data/100715.fcs")
f

colnames(f)  # Channel labels

# Expression values for each event
E <- exprs(f)
dim(E)

# Let's look at the first 10 rows of E
E[1:10,]

# This provides the keyword information stored in the file
f@description

# Specific keyword
f@description$P9DISPLAY

#Access paramter info such as a range of expression values for range of expression values
# here we are doing it for forward-scatter light
f@parameters@data
f@parameters@data[1, c("minRange", "maxRange")]

```

Let's plot this scheet which is part of the package **flowViz**
```{r}

# Cool, so 'f' is an object for a single flowFrame object, holding a single FCS file
plot(f, 
     c("FSC-A", "SSC-A"), 
     #xlim = c(), 
     ylim = c(0,5000), 
     smooth = FALSE)

# SSC-A is the third parameter (P3) and the meta-date will tell us to view it on log scale
colnames(f)[3]

f@description$"P3DISPLAY"
```

Let's make an obejct out of a set of FCS files using the flowSet command. Note, it's kind of weird if you have FCS files from different machines, so make sure that your directory is from a single machine

```{r}
fs <- read.flowSet(path = './tutorial/data', pattern = ".fcs")
fs

# Let's look at the metadata
sampleNames(fs)
length(fs)

# flowSet objects are sort of like lists in that you can access the first frame by index or name
fs[["100715.fcs"]]
fs[[1]]

# Now we can use fsApply to get event counts for all samples
nrow(fs[[1]])
fsApply(fs, nrow)

# fsApply can be used to extract the TUBE NAME keyword. You will want this for identifying control or stained tubes or various treatments, whatever
# This depends on you ACTUALLY SPECIFYING THIS during your FC run with Christiane. In this example, each tube is a different file, but those generated at the IU Core Facility will have several tubes associated with an individiual FCS file
fsApply(fs, function(f) f@description$"TUBE NAME")
```

The plot command from flowVis will only handle one FCS file at a time, but you can subset the flowset with brackets.

```{r}
par(mfrow = c(1,3))

plot(fs[[1]], 
     c("FSC-A", "SSC-A"), 
     #xlim = c(), 
     ylim = c(0,5000), 
     smooth = FALSE)

plot(fs[[2]], 
     c("FSC-A", "SSC-A"), 
     #xlim = c(), 
     ylim = c(0,10000), 
     smooth = FALSE)

plot(fs[[3]], 
     c("FSC-A", "SSC-A"), 
     #xlim = c(), 
     ylim = c(0,10000), 
     smooth = FALSE)

```

#Module 3: Preprocessing and Quality Assurance of FCM Data
(PDF)[http://bioinformatics.ca//files/public/FACS_2013_Module3_Lecture_v2.pdf]; 

FCM data requires a pipeline for processing. This means that we must first **compensate** the data appropriately, then construct an scripting algorithm to objectively remove debris, doublets, and margin events. Next, we will discuss transformation strategies for the data, remove margin events, and proceed with gating.

## 3.1 Preprocessing
### 3.1.1 Compensation with the flowCore package for a single sample
```{r}
# Let's first look at the **compensate** help menu
?compensate

M <- f@description$'SPILL'
M

# We can compensate by supplying the flowFrame object and the compensation matrix
f.comp <- compensate(f, M)
summary(f)
summary(f.comp)

```

### 3.1.2 Removing Margin Events

```{r}
length(fs)
fs[[2]]

# Let's look at the scatter channels first. There are margin events at the upper end of the FSC-A channel

plot(fs[[2]], c("FSC-A", "SSC-A"),
     ylim = c(0,5000),
     smooth = FALSE)

# We can add a rough cut off point for the margin events
abline(v = 250000, col = "blue", lwd = 3, lty = "dashed")

# First, select the margin events
margin.cells <- which(exprs(fs[[1]])[,"FSC-A"] >= 250000)
length(margin.cells)
nrow(fs[[1]])

# Calculate the percentage of cells on the FSC-A margin:
margin.perc <- 100 * length(margin.cells)/nrow(fs[[1]])
margin.perc
```

### 3.1.3 Visualize Margin Events
```{r}
f <- fs[[2]]
A <- exprs(f)[,c("FSC-A", "SSC-A")]

points(A[margin.cells,], pch = ".", col = "red", cex = 3)
legend('top', legend = paste("Margin Events:",margin.perc, "%"),
       col = "red", pch = 19)

f.clean.margin <- f[-margin.cells]
nrow(f.clean.margin)
```

### 3.1.3 Transformations

**Why transform your data?**

**log** cannot handle negative values
**biexponential** these make use of ***arcsinh*** and ***logicle*** transformations. ***arcsinh*** transform is used for data with few or no negative expression values after compensations. ***logicle*** transform is similar to arcsinh but with greater parameterization options.

Let's look at a simple example:
```{r}
# Simple transformations
a <- c(1,10,100,500,1000)
log10(a)
asinh(a)

lgcl <- logicleTransform()
print (lgcl(a))

```

Now, with our cleaned data, let's try some different transformations. 
```{r}
vals <- exprs(f.clean.margin)[,"R780-A"]
vals[1:4]

# Set up plot region

par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,1,0))
plot(density(vals), 
     xlim = c(0,20000),
     main = "Untransformed CD3 values")
plot(density(log10(vals), na.rm = TRUE), main = "Log Transform")
plot(density(asinh(vals)), main = "Asinh")
plot(density(lgcl(vals)), main = "Logicle")

```

### 3.1.4 Preprocessing for flowSets
```{r}
# we can also do this for the whole flowSet object
fs.comp <- fsApply(fs,function(frame){
  #extract compensation matrix from keywords
  comp <- keyword(frame)$`SPILL`
  new_frame <- compensate(frame,comp)
  new_frame
})

fs.comp
```

### 3.1.5 Planning the removal algorithm

## 3.2 Quality Assurance using the flowQ package

Before continuing with the data analysis, run some QC on the data to flag any suspicious samples. QC may include:

**High margin cell counts** which may indicate a technical issue with aquistion
**Viable cell count** 
**Density check** 

#4. 1D Staitc gating
#5. 1D Dynamic gating
#6. Clustering and Additional FCM tools
#7. Example: Analyzing DNA/RNA content in bacterial cells
##7.1 Data collection and methodology
##7.2 Explore data
##7.3 Create gates for analysis